Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2022-02-15 Manuscript/SS_GAM.tex
Update Section Headers to Latex

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\label{sec:intro}

@@

\hypertarget{bayesian-hierarchical-additive-models-bhams}{%
\section{Bayesian Hierarchical Additive Models
(BHAMs)}\label{bayesian-hierarchical-additive-models-bhams}}

\label{sec:BHAM}

@@

\hypertarget{spike-and-slab-spline-priors}{%
\subsection{Spike-and-Slab Spline
Priors}\label{spike-and-slab-spline-priors}}

@@

\hypertarget{two-part-spike-and-slab-lasso-prior}{%
\subsubsection{Two-part Spike-and-Slab Lasso
Prior}\label{two-part-spike-and-slab-lasso-prior}}

@@

\hypertarget{other-priors}{%
\subsubsection{Other Priors}\label{other-priors}}

@@

\hypertarget{algorithms-for-fitting-bhams}{%
\subsection{Algorithms for Fitting
BHAMs}\label{algorithms-for-fitting-bhams}}

@@

\hypertarget{em-algorithms}{%
\subsubsection{EM algorithms}\label{em-algorithms}}

@@

\hypertarget{em-coordinate-descent}{%
\subsubsection{EM-Coordinate Descent}\label{em-coordinate-descent}}

\label{sec:EMCD}

@@

\hypertarget{em-iwls}{%
\subsubsection{EM-IWLS}\label{em-iwls}}

Similar to the EM-CD algorithm, the EM-IWLS algorithm is an iterative
EM-based algorithm where the iterative weighted least squares algorithm
is used to find the estimate of \(\boldsymbol{\beta}, \phi\) that
maximizes \(E(Q_1)\). The iterative weighted least squares algorithm was
originally proposed to fit the classical generalized linear models, and
generalized to fit some Bayesian hierarchical models.\citep{Gelman2013}
Yi and Ma \citep{Yi2012} formulated Student's t-distribution and double
exponential distribution as hierarchical normal distributions such that
generalized linear models with shrinkage priors can be easily fitted
using IWLS in combination with EM algorithm. In this work, we adapt the
EM-IWLS paradigm to fit BHAM with spike-and-slab spline prior .

A double exponential prior, \(\beta|S \sim DE(0, S)\) can be formulated
as a hierarchical normal prior with unknown variance \(\tau^2\)
integrated out: \begin{align*}
  \beta|\tau^2 &\sim N(0, \tau^2)\\
  \tau^2|S & \sim Gamma(1, 1/(2S^2)), 
\end{align*} For the mixture double exponential priors, we can define
the scale parameter \(S = (1-\gamma)s_0 + \gamma s_1\) following
Equation (\ref{eq:ssl}). The change in the prior formulation in turn
leads to the change in the log posterior density function, as \(Q_1\)
needs to account for the hyperprior of \(\tau^2\):
\begin{equation}\label{eq:Q1_IWLS}
Q_1(\boldsymbol{\beta}, \phi) = \log f(\textbf{y}|\boldsymbol{\beta}, \phi) + \sum\limits_{j=1}^p\left[\log f(\beta_j|{\tau}^2_j) + \log f({\tau}^2_j| S_j)+\sum\limits_{k=1}^{K_j} \{\log f(\beta^{*}_{jk}|{\tau^{*}}^2_{jk})+\log f({\tau^*}^2_{jk}| S^*_j)\}\right].
\end{equation} Since \(\boldsymbol{\tau}^2\) are not of our primary
interest, we treat them as the ``missing'' data in addition to the
latent indicators \(\boldsymbol{\gamma}\), and hence construct the
expectation
\(E_{\boldsymbol{\gamma}, \boldsymbol{\tau}^2|\Theta^{(t-1)}}(Q_1)\) in
the E-step. To note, unlike the same latent indicator \(\gamma^*_j\)
which is shared by the coefficients of the non-linear terms
\(\beta^*_{jk}\) for \(k = 1, \dots, K_j\) , \(\tau^2_{jk}\) is
coefficient specific for \(\beta^*_{jk}\).
\(E({S_j}^{-1}|\beta_j, s_0, s_1), E({S^*}^{-1}_j|\boldsymbol{\beta}_j^*, s_0, s_1), E({\tau}^2_{j}|S_j, \beta_j) \text{ and } E({\tau^*}^2_{jk}|S_j^*, \beta^*_{jk})\)
needs to be calculated to formulate \(E(Q_1)\). As neither
\(E({S_j}^{-1}|\beta_j, s_0, s_1)\) nor
\(E({S^*}^{-1}_j|\boldsymbol{\beta}_j^*, s_0, s_1)\) depends on
\(\tau^2\)s, they can be derived using Equation (\ref{eq:exp_scale}). On
the other hand, \(\tau^{2}\), following gamma distributions, is a
conjugate prior for the normal variance, and the conditional posterior
density of \(\tau^{-2}\) is an inverse Gaussian distribution.
\(E({\tau}^{-2}_{j})\) and \(E({\tau^*}^{-2}_{jk})\) are calculated
using the closed form equation \begin{align*}
 E({\tau}^{-2}_{j}|S_j, \beta_j) ={S_j}^{-1}/|\beta_j| \qquad E({\tau^*}^{-2}_{jk}|S_j^*, \beta^*_{jk})={S_j^*}^{-1}/|\beta^*_{jk}|,
\end{align*} where \(S_j\) and \(S_j^*\) are replaced by the expectation
and \(\beta\)s are replaced with \(\beta^{(t-1)}\). With simplification
(up to constant additive terms), we have
\begin{equation}\label{eq:EQ1_IWLS}
E(Q_1) = \log f(\textbf{y}|\boldsymbol{\beta}, \phi) - \sum\limits_{j=1}^p\left[ {2E({\tau_j}^{-2})}{\beta_j}^2 +\sum\limits_{k=1}^{K_j} {2E({\tau_{jk}^*}^{-2})}{\beta_{jk}^*}^2\right].
\end{equation} \(2E({\tau}^{-2})\beta^2\) can be seen as the kernel of a
normal density with mean 0 and variance \(E(\tau^{2})\), and we can
formulate the coefficients \(\boldsymbol{\beta}\) as a multivariate
normal distribution with means \(\boldsymbol{0}\) and variance
covariance matrix \(\boldsymbol{\Sigma}_{\tau^2}\), where
\(\boldsymbol{\Sigma}_{\tau^2}\) is a diagonal matrix with
\(E(\tau^2)\)s on the diagonal, \[
\boldsymbol{\beta }\sim \text{MVN}(0, \boldsymbol{\Sigma}_{\tau^2}).
\]

Meanwhile, following the classical IWLS, we can approximate the
generalized model likelihood at each iteration with a weighted normal
likelihood: \[
f(\textbf{y}|\boldsymbol{\beta}, \phi) \approx \text{MVN}(\textbf{z}|\boldsymbol{X} \boldsymbol{\beta}, \phi\boldsymbol{\Sigma })
\] where the `normal response' \(z_i\) and `weight' \(w_i\) are called
the pseudo-response and pseudo-weight respectively. The pseudo-response
and the pseudo-weight are calculated by \[
\begin{aligned}
z_i &= \hat\eta_i - \frac{L^{'}(y_i|\hat\eta_i)}{L^{''}(y_i|\hat\eta_i)}& w_i &= - L^{''}(y_i|\hat\eta_i),
\end{aligned}
\] where \(\hat\eta_i = (\boldsymbol{X} {\hat{\boldsymbol{\beta}}})_i\),
\(L^{'}(y_i|\hat\eta_i, \hat \phi)\) and
\(L^{''}(y_i|\hat\eta_i, \hat \phi)\) are the first and second
derivative of the log density,
\(\log f(\textbf{y}_i|\boldsymbol{\beta}, \phi)\) with respect to
\(\eta_i\).

With
\(\boldsymbol{z}\sim \text{MVN}(\boldsymbol{X} \boldsymbol{\beta}, \phi \boldsymbol{\Sigma})\)
and
\(\boldsymbol{\beta }\sim \text{MVN}(0, \phi \boldsymbol{\Sigma}_{\tau^2})\),
we can augment the two multivariate normal distributions and update the
estimates for \(\boldsymbol{\beta}\) and \(\phi\) via least squares in
each iteration of the EM algorithm. We create the augmented response,
augmented data, and augmented variance-covariance matrix following
\begin{align*}
& \boldsymbol{z}_* = \begin{bmatrix} \boldsymbol{z}\\ \boldsymbol{0}\end{bmatrix} &&
  \boldsymbol{X}_* = \begin{bmatrix} \boldsymbol{X} \\ \boldsymbol{I} \end{bmatrix} &&
  \boldsymbol{\Sigma}_* = \begin{bmatrix} \boldsymbol{\Sigma }& \boldsymbol{0}  \\ \boldsymbol{0} & \boldsymbol{\Sigma}_{\tau^2}/\phi \end{bmatrix}, &
\end{align*} such that \[
\boldsymbol{z}_* \sim \text{MVN}(\boldsymbol{X}_* \boldsymbol{\beta }, \phi \Sigma_*).
\] Using the least squares estimators to update \(\boldsymbol{\beta}\)
and \(\phi\), we have \begin{align*}
& \boldsymbol{\beta}^{(t)} = (\boldsymbol{X}_*^T \boldsymbol{\Sigma}^{-1} \boldsymbol{X}_*)^{-1}\boldsymbol{X}_*^T \boldsymbol{\Sigma}^{-1} \boldsymbol{z}_* && \phi^{(t)} = \frac{1}{n}(\boldsymbol{z}_*-X_*\boldsymbol{\beta}^{(t)})^T\boldsymbol{\Sigma}^{-1}(\boldsymbol{z}_*-X_*\boldsymbol{\beta}^{(t)}).&
\end{align*} To note, the variance-covariance matrix of the coefficient
estimates variance-covariance matrix can be derived in the EM-IWLS
algorithm and in turn can be used for statistical inferences, \[
  \text{Var}(\boldsymbol{\beta}^{(t)}) = (\boldsymbol{X}_*^T\boldsymbol{\Sigma}^{-1} \boldsymbol{X}_*)^{-1}\phi^{(t)}.
\]

Totally, the proposed EM-IWLS algorithm is summarized as follows:

  E-step: calculate \(E(\gamma_{j})\), \(E(\gamma^*_{j})\) and
  \(E(\tau^{-2}_{j})\), \(E({\tau^*}^{-2}_{jk})\) with the estimates
  \(\Theta^{(t-1)}\) from the previous iteration

  \begin{enumerate}
  \def\labelenumii{\alph{enumii})}
  \tightlist
  \item
    Based on the current value of \(\beta\), calculate the pseudo-data
    \(z_i^{(t)}\) and the pseudo-weights \(w_i^{(t)}\)
  \item
    Update \(\boldsymbol{\beta}^{(t)}\) by runing the augmented weighted
    least squared
  \item
    If \(\phi\) is present, update \(\phi\)
  \end{enumerate}
\end{enumerate}

Similar to EM-CD, we assess convergence by the criterion,
\(|d^{(t)}-d^{(t-1)}|/(0.1+|d^{(t)}|)<\epsilon\), where \(\epsilon\) is
a small value (say \(10^{-5}\)).

\hypertarget{selecting-optimal-scale-values}{%
\subsection{Selecting Optimal Scale
Values}\label{selecting-optimal-scale-values}}

\label{sec:tune} Our proposed models, BHAM, require two preset scale
parameters (\(s_0\), \(s_1\)). Hence, we need to find the optimal values
for the scale parameters such that the model reaches its best prediction
performance regarding a criteria of preference. This would be achieved
by constructing a two dimensional grid, consists of different pairs of
(\(s_0\), \(s_1\)) value. However, previous research suggested the value
of slab scale \(s_1\) have less impact on the final model and is
recommended to be set as a generally large value, e.g.~\(s_1 = 1\), that
provides no or weak shrinkage. \citep{Rockova2018} As a result, we focus
on examining different values of spike scale \(s_0\). Instead of the 2-D
grid, We consider a sequence of \(L\) decreasing values
\(\{s_0^l\}: 0 < s_0^1 < s_0^2 < \dots < s_0^L < s_1\). Increasing the
spike scale \(s_0\) tends to include more non-zero coefficients in the
model. A measure of preference calculated with cross-validations (CV),
e.g.~deviance, area under the curve (AUC), mean squared error, can be
used to facilitate the selection of a final model. The procedure is
similar to the Lasso implemented in the widely used R package
\texttt{glmnet}, which quickly fits Lasso models over a list of values
of regularization parameters \(\lambda\) , giving a sequence of models
for users to choose from.

\hypertarget{simulation-study}{%
\section{Simulation Study}\label{simulation-study}}

\label{sec:sim}

@@

\hypertarget{monte-carlo-simulation-study}{%
\subsection{Monte Carlo Simulation
Study}\label{monte-carlo-simulation-study}}

@@

@@

@@

\hypertarget{metabolomics-data-analysis}{%
\section{Metabolomics Data Analysis}\label{metabolomics-data-analysis}}

\label{sec:real_data}

@@

\hypertarget{emory-cardiovascular-biobank}{%
\subsection{Emory Cardiovascular
Biobank}\label{emory-cardiovascular-biobank}}

\label{sec:ECB} We use the proposed models BHAM to analyze a metabolic
dataset from a recently published research \citep{Mehta2020} studying
plasma metabolomic profile on the three-year all-cause mortality among
patients undergoing cardiac catheterization. The dataset is publicly
available via \textit{Dryad} \citep{Mehta2020_data}. It contains in
total of 776 subjects from two cohorts. As there is a large number of
non-overlapping features among the two cohorts, we use the cohort with
larger sample size (N=454). There are initially 6796 features in the
dataset, which is too large to be practically meaningful to analyze.
Hence, we perform a univariate screening procedure on the features, via
GAM implemented in \texttt{mgcv}, and choose the the top 200 features
with smallest p-values. We use 5-knot spline additive models for binary
outcome using two different models, the proposed BHAM and the SB-GAM.
10-Fold CV are used to choose the optimal tuning parameters of each
framework with respect to the default selection criterion implemented in
the software. Out-of-bag samples are used for prediction performance
evaluation, where deviance, AUC, Brier score defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and
misclassication error defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)\) are
calculated. BHAM-CD obtains superior AUC, Brier score, and
misclassification error in the out-of-bag samples compared to SB-GAM
(see Table \ref{tab:ECB_res}).

@@

\hypertarget{weight-loss-maintenance-cohort}{%
\subsection{Weight Loss Maintenance
Cohort}\label{weight-loss-maintenance-cohort}}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\label{sec:concl} In the paper, we described a novel generalized
additive model using Bayesian hierarchical priors, particularly the
proposed spike-and-slab spline prior for bi-level functional selection.
Meanwhile, we introduced two optimization based algorithms for model
fitting. The algorithms can be easily scale up to address
high-dimensional data analysis in a computational efficient manner. Via
simulations, we demonstrated that the proposed model provides as good,
if not better, prediction performance compared to some state-of-the-art
non-linear modeling devices.

@@

\newpage

\begin{table}[ht]
\centering
\begin{tabular}{ccccccc}
  \hline
P & BHAM-IWLS & BHAM-CD & COSSO & Adaptive COSSO & mgcv & SB-GAM \\ 
  \hline
  4 & 0.90 (0.01) & 0.90 (0.01) & 0.75 (0.03) & 0.71 (0.13) & 0.90 (0.01) & 0.82 (0.04) \\ 
   10 & 0.90 (0.01) & 0.88 (0.01) & 0.67 (0.15) & 0.76 (0.03) & 0.90 (0.01) & 0.82 (0.04) \\ 
   50 & 0.88 (0.01) & 0.80 (0.04) & 0.43 (0.17) & 0.57 (0.19) & 0.86 (0.02) & 0.82 (0.04) \\ 
  100 & 0.80 (0.07) & 0.73 (0.06) & 0.41 (0.19) & 0.51 (0.22) & - & 0.81 (0.04) \\ 
  200 & 0.72 (0.10) & 0.77 (0.02) & 0.33 (0.15) & 0.44 (0.19) & - & 0.82 (0.04) \\ 
   \hline
\end{tabular}
\caption{The average and standard deviation of the out-of-sample $R^2$ measure for Gaussian outcomes over 50 iterations. The models of comparison include the proposed Bayesian hierarchical additive model (BHAM) fitted with Iterative Weighted Least Square (BHAM-IWLS) and Coordinate Descent (BHAM-CD) algorithms, component selection and smoothing operator (COSSO), adaptive COSSO, mgcv and sparse Bayesian generalized additive model (SB-GAM). mgcv doesn't provide estimation whe number of parameters exceeds sample size i.e. p = 100, 200.} 
\label{tab:gaus}
\end{table}

>>>>

\section{Introduction}
\label{sec:intro}

@@

\section{Bayesian Hierarchical Additive Models (BHAMs)}
\label{sec:BHAM}

@@

\subsection{Spike-and-Slab Spline Priors}

@@

\subsubsection{Two-part Spike-and-Slab Lasso Prior}

@@

\subsubsection{Other Priors}

@@

\subsection{Algorithms for Fitting BHAMs}

@@

\subsubsection{EM algorithms}

@@

\subsubsection{EM-Coordinate Descent}
\label{sec:EMCD}

@@

\subsection{Selecting Optimal Scale Values}
\label{sec:tune}

Our proposed models, BHAM, require two preset scale parameters (\(s_0\),
\(s_1\)). Hence, we need to find the optimal values for the scale
parameters such that the model reaches its best prediction performance
regarding a criteria of preference. This would be achieved by
constructing a two dimensional grid, consists of different pairs of
(\(s_0\), \(s_1\)) value. However, previous research suggested the value
of slab scale \(s_1\) have less impact on the final model and is
recommended to be set as a generally large value, e.g.~\(s_1 = 1\), that
provides no or weak shrinkage. \citep{Rockova2018} As a result, we focus
on examining different values of spike scale \(s_0\). Instead of the 2-D
grid, We consider a sequence of \(L\) decreasing values
\(\{s_0^l\}: 0 < s_0^1 < s_0^2 < \dots < s_0^L < s_1\). Increasing the
spike scale \(s_0\) tends to include more non-zero coefficients in the
model. A measure of preference calculated with cross-validations (CV),
e.g.~deviance, area under the curve (AUC), mean squared error, can be
used to facilitate the selection of a final model. The procedure is
similar to the Lasso implemented in the widely used R package
\texttt{glmnet}, which quickly fits Lasso models over a list of values
of regularization parameters \(\lambda\) , giving a sequence of models
for users to choose from.

\section{Simulation Study}
\label{sec:sim}

@@

\subsection{Monte Carlo Simulation Study}

@@

\subsection{Simulation Results}
\subsubsection{Prediction Performance}

@@

\subsubsection{Variable Selection Performance}

@@

\section{Metabolomics Data Analysis}
\label{sec:real_data}

@@

\subsection{Emory Cardiovascular Biobank}
\label{sec:ECB}

We use the proposed models BHAM to analyze a metabolic dataset from a
recently published research \citep{Mehta2020} studying plasma
metabolomic profile on the three-year all-cause mortality among patients
undergoing cardiac catheterization. The dataset is publicly available
via \textit{Dryad} \citep{Mehta2020_data}. It contains in total of 776
subjects from two cohorts. As there is a large number of non-overlapping
features among the two cohorts, we use the cohort with larger sample
size (N=454). There are initially 6796 features in the dataset, which is
too large to be practically meaningful to analyze. Hence, we perform a
univariate screening procedure on the features, via GAM implemented in
\texttt{mgcv}, and choose the the top 200 features with smallest
p-values. We use 5-knot spline additive models for binary outcome using
two different models, the proposed BHAM and the SB-GAM. 10-Fold CV are
used to choose the optimal tuning parameters of each framework with
respect to the default selection criterion implemented in the software.
Out-of-bag samples are used for prediction performance evaluation, where
deviance, AUC, Brier score defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and
misclassication error defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)\) are
calculated. BHAM-CD obtains superior AUC, Brier score, and
misclassification error in the out-of-bag samples compared to SB-GAM
(see Table \ref{tab:ECB_res}).

@@

\subsection{Weight Loss Maintenance Cohort}

@@

\section{Discussion}
\label{sec:concl}

In the paper, we described a novel generalized additive model using
Bayesian hierarchical priors, particularly the proposed spike-and-slab
spline prior for bi-level functional selection. Meanwhile, we introduced
two optimization based algorithms for model fitting. The algorithms can
be easily scale up to address high-dimensional data analysis in a
computational efficient manner. Via simulations, we demonstrated that
the proposed model provides as good, if not better, prediction
performance compared to some state-of-the-art non-linear modeling
devices.

@@

\clearpage

\begin{table}[ht]
\centering
\begin{tabular}{ccccccc}
  \hline
P & BHAM-IWLS & BHAM-CD & COSSO & Adaptive COSSO & mgcv & SB-GAM \\ 
  \hline
  4 & 0.90 (0.01) & 0.90 (0.01) & 0.75 (0.03) & 0.71 (0.13) & 0.90 (0.01) & 0.82 (0.04) \\ 
   10 & 0.90 (0.01) & 0.88 (0.01) & 0.67 (0.15) & 0.76 (0.03) & 0.90 (0.01) & 0.82 (0.04) \\ 
   50 & 0.88 (0.01) & 0.80 (0.04) & 0.43 (0.17) & 0.57 (0.19) & 0.86 (0.02) & 0.82 (0.04) \\ 
  100 & 0.80 (0.07) & 0.73 (0.06) & 0.41 (0.19) & 0.51 (0.22) & - & 0.81 (0.04) \\ 
  200 & 0.72 (0.10) & 0.77 (0.02) & 0.33 (0.15) & 0.44 (0.19) & - & 0.82 (0.04) \\ 
   \hline
\end{tabular}
\caption{The average and standard deviation of the out-of-sample $R^2$ measure for Gaussian outcomes over 50 iterations. The models of comparison include the proposed Bayesian hierarchical additive model (BHAM) fitted with Iterative Weighted Least Square (BHAM-IWLS) and Coordinate Descent (BHAM-CD) algorithms, component selection and smoothing operator (COSSO), adaptive COSSO, mgcv and sparse Bayesian generalized additive model (SB-GAM). mgcv doesn't provide estimation whe number of parameters exceeds sample size i.e. p = 100, 200.} 
\label{tab:gaus}
\end{table}
\clearpage
